---
title: "R Intro Spatial Stats"
author: "Jakob Dambon"
date: "2023-05-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Example

The following simple R example is derived from the lecture notes of [STA330: Modeling Dependent Data](http://user.math.uzh.ch/furrer/download/sta330/script_sta330.pdf) by Prof. Reinhard Rurrer, University of Zurich.

## Data

```{r}
library(fields)

?ChicagoO3
# 20 ozone measurements in 1987
y <- ChicagoO3$y
summary(y)
```

```{r}
hist(y)
```


## Locations

```{r}
# First in Latitude and Longitude (good for visualization)
summary(ChicagoO3$lon.lat)
```


```{r}
# Now in approx Cartesian (in miles)
summary(ChicagoO3$x)
plot(ChicagoO3$x)
```


## Visualization of Data

```{r}
# combine data and locations
library(sp)
sp_ozone <- SpatialPointsDataFrame(
  ChicagoO3$lon.lat, 
  data.frame(ozone = ChicagoO3$y), 
  # projections
  proj4string = CRS("+proj=longlat +datum=WGS84")
)

str(sp_ozone)
```

```{r}
# first plot. Note that the dimensions are not distorted
spplot(sp_ozone, colorkey = TRUE)
```


```{r}
# very powerful mapping tool (similar syntax to ggplot2)
library(tmap)
# interactive mode
tmap_mode("view")

tm_shape(sp_ozone) + 
  tm_symbols(col = "ozone")
```


## Modeling using a Gaussian Process

We need the following components:

### Distance Matrix: 

```{r}
# using the Cartesian locations
distmat <- as.matrix(dist(ChicagoO3$x))
# distance matrix heat map. Note that the diagonal is left bottom to top right
spam::image(distmat)
```


### Choice of a Covariance function

```{r}
x <- seq(0, 5, by = 0.1)
l_cf <- list(
  "exponential" = spam::cov.exp, 
  "spherical" = spam::cov.sph, 
  "Matern3/2" = spam::cov.mat32
)

matplot(
  x, 
  sapply(
    l_cf, 
    function(cf) 
      # theta: 1 range, 2, variance, 3 nugget
      do.call(cf, args = list(h = x, theta = c(1, 1, 0)))
  ),
  type = "l",
  ylab = "c(d)",
  xlab = "d",
  main = "Some Covariance Functions"
)
legend(
  "topright", 
  legend = names(l_cf), 
  col = seq_along(l_cf), 
  lty = 1, lwd = 1.2
)
```


### Maximum Likelihood Estimation

```{r}
# we will fit a model without covariates
X <- matrix(1, length(ChicagoO3$y), 1)

# y ~ beta0 + Z(s), where Z(s) is a Gaussian Process 
(pars <- spam::mle(
  # data & distance information
  y, X, distmat, cov.sph, 
  # initial values of optimization
  beta0 = 0, theta0 = c(2, 2),
  # Bounds (covariance parameters need to be positive)
  thetalower = c(.01,.01), thetaupper = Inf)
)
```


### What we derive:

```{r}
# estimated mean (beta0)
pars$par[1]
# compared to empirical mean 
mean(y)
```

```{r}
# estimated range
pars$par[2]
# is equal to the initial value. Probably too few observations to estimate.
```

```{r}
# estimated variance
pars$par[3]
# compared to empirical variance
var(y)

# most of the variance can be explained by the spatial distribution
pars$par[3]/var(y)
```

### Further Reading


Please check out the vignette of the `gstat` package (`vignette("gstat", package = "gstat")`) for a further example. 


